---
title: "Make Circos Plots"
author: "Seidu Adams<br><small>Fernando's Lab<br>Department of Animal Science<br>University of Nebraska -Lincoln</small>"
date: "<small>`r Sys.Date()`</small>"
output:
  html_document:
    toc: yes
    toc_depth: 3
    number_sections: true
    toc_float: 
      collapsed: true
    code_folding: show
    theme: cerulean
    self_contained: true
  pdf_document: 
    latex_engine: xelatex
  word_document: default  
---


Now everything is ready to start creating Chord diagram in R. First let’s install and load the circlize package as shown below:

```{r}
#install and load the package
#install.packages("circlize")
library(circlize)
```


Then we need to read the file into R by following the command below. Make sure the file is in CSV format and change the location in the command below. 

```{r}
#setwd("E:/ORAL_RUMEN_COMPARATIVE_ANALYSIS/FINAL_ORAL_RUMEN_2022YEAR")
#read the csv file
data <- read.csv("P400.468.family.csv", row.names = 1)
```



After that, we need to convert our table to a matrix

```{r}
#convert the table to a martix
data <- as.matrix(data)
```


We can immediately plot a chord diagram by using the chordDiagram() function.

```{r}
#create a chord diagram
chordDiagram(data)
```


R created the plot in the default setting. So we need to adjust according to our preferences.


Since we have many columns and rows, the labels are on top of each other. To solve this, we need to plot the chord diagram without labels and axis and then adding them separately as shown in the commands below:

```{r}
jpeg("P400.468.jpg", height = 15, width = 15, units = 'in', res = 600)
#create a chord diagram but without labeling 
chordDiagram(data, annotationTrack = "grid", preAllocateTracks = 1)

#add the labels and axis
circos.trackPlotRegion(track.index = 2, panel.fun = function(x, y) {
  xlim = get.cell.meta.data("xlim")
  ylim = get.cell.meta.data("ylim")
  sector.name = get.cell.meta.data("sector.index")
  
#print labels 
circos.text(mean(xlim), ylim[1] + 2.5, sector.name, 
              facing = "clockwise", niceFacing = TRUE, adj = c(0, 0.5), cex=0.6)
  
#print axis
circos.axis(h = "top", labels.cex = 0.5, major.tick.percentage = 0.2, 
              sector.index = sector.name, track.index = 2)
}, bg.border = NA)
```





Then we need to read the file into R by following the command below. Make sure the file is in CSV format and change the location in the command below. 

```{r}
#setwd("E:/ORAL_RUMEN_COMPARATIVE_ANALYSIS/FINAL_ORAL_RUMEN_2022YEAR")
#read the csv file
data <- read.csv("P295.365.family.csv", row.names = 1)
```



After that, we need to convert our table to a matrix

```{r}
#convert the table to a martix
data <- as.matrix(data)
```


We can immediately plot a chord diagram by using the chordDiagram() function.

```{r}
#create a chord diagram
chordDiagram(data)
```


R created the plot in the default setting. So we need to adjust according to our preferences.


Since we have many columns and rows, the labels are on top of each other. To solve this, we need to plot the chord diagram without labels and axis and then adding them separately as shown in the commands below:

```{r}
jpeg("P295.365.jpg", height = 15, width = 15, units = 'in', res = 600)
#create a chord diagram but without labeling 
chordDiagram(data, annotationTrack = "grid", preAllocateTracks = 1)

#add the labels and axis
circos.trackPlotRegion(track.index = 2, panel.fun = function(x, y) {
  xlim = get.cell.meta.data("xlim")
  ylim = get.cell.meta.data("ylim")
  sector.name = get.cell.meta.data("sector.index")
  
#print labels 
circos.text(mean(xlim), ylim[1] + 2.5, sector.name, 
              facing = "clockwise", niceFacing = TRUE, adj = c(0, 0.5), cex=0.6)
  
#print axis
circos.axis(h = "top", labels.cex = 0.5, major.tick.percentage = 0.2, 
              sector.index = sector.name, track.index = 2)
}, bg.border = NA)
```





Then we need to read the file into R by following the command below. Make sure the file is in CSV format and change the location in the command below. 

```{r}
#setwd("E:/ORAL_RUMEN_COMPARATIVE_ANALYSIS/FINAL_ORAL_RUMEN_2022YEAR")
#read the csv file
data <- read.csv("P170.240.family.csv", row.names = 1)
```



After that, we need to convert our table to a matrix

```{r}
#convert the table to a martix
data <- as.matrix(data)
```


We can immediately plot a chord diagram by using the chordDiagram() function.

```{r}
#create a chord diagram
chordDiagram(data)
```


R created the plot in the default setting. So we need to adjust according to our preferences.


Since we have many columns and rows, the labels are on top of each other. To solve this, we need to plot the chord diagram without labels and axis and then adding them separately as shown in the commands below:

```{r}
jpeg("P170.240.jpg", height = 15, width = 15, units = 'in', res = 600)
#create a chord diagram but without labeling 
chordDiagram(data, annotationTrack = "grid", preAllocateTracks = 1)

#add the labels and axis
circos.trackPlotRegion(track.index = 2, panel.fun = function(x, y) {
  xlim = get.cell.meta.data("xlim")
  ylim = get.cell.meta.data("ylim")
  sector.name = get.cell.meta.data("sector.index")
  
#print labels 
circos.text(mean(xlim), ylim[1] + 2.5, sector.name, 
              facing = "clockwise", niceFacing = TRUE, adj = c(0, 0.5), cex=0.6)
  
#print axis
circos.axis(h = "top", labels.cex = 0.5, major.tick.percentage = 0.2, 
              sector.index = sector.name, track.index = 2)
}, bg.border = NA)
```




Then we need to read the file into R by following the command below. Make sure the file is in CSV format and change the location in the command below. 

```{r}
#setwd("E:/ORAL_RUMEN_COMPARATIVE_ANALYSIS/FINAL_ORAL_RUMEN_2022YEAR")
#read the csv file
data <- read.csv("P50.120_Family.csv", row.names = 1)
```



After that, we need to convert our table to a matrix

```{r}
#convert the table to a martix
data <- as.matrix(data)
```


We can immediately plot a chord diagram by using the chordDiagram() function.

```{r}
#create a chord diagram
chordDiagram(data)
```


R created the plot in the default setting. So we need to adjust according to our preferences.


Since we have many columns and rows, the labels are on top of each other. To solve this, we need to plot the chord diagram without labels and axis and then adding them separately as shown in the commands below:

```{r}
jpeg("P50.120.jpg", height = 15, width = 15, units = 'in', res = 600)
#create a chord diagram but without labeling 
chordDiagram(data, annotationTrack = "grid", preAllocateTracks = 1)

#add the labels and axis
circos.trackPlotRegion(track.index = 2, panel.fun = function(x, y) {
  xlim = get.cell.meta.data("xlim")
  ylim = get.cell.meta.data("ylim")
  sector.name = get.cell.meta.data("sector.index")
  
#print labels 
circos.text(mean(xlim), ylim[1] + 2.5, sector.name, 
              facing = "clockwise", niceFacing = TRUE, adj = c(0, 0.5), cex=0.6)
  
#print axis
circos.axis(h = "top", labels.cex = 0.5, major.tick.percentage = 0.2, 
              sector.index = sector.name, track.index = 2)
}, bg.border = NA)
```










You do not need to change anything in the commands above. The argument that changes the labels’ angel is facing = “clockwise”. Now we get a beautiful plot with readable labels.



You will find the saved images in your directory. One thing you might noticed is that the color is changing each time you create a chord diagram. This is because R assigns colors randomly.


You may want to assign specific colors for some groups to make the figure more informative and also easy to read. To do that, we need to assign a color for each group as shown below. I put the whole previous codes in addition to the coloring command (colored in red).

```{r}
jpeg("Circos2_Combined.jpg", height = 10, width = 15, units = 'in', res = 600)
#Color the links

#read the csv file
data <- read.csv("DAT.csv", row.names = 1)

#convert the table to a martix
data <- as.matrix(data)

#assign color to each group of strains
col = c(Litter_1="#330033", Sow_1="#330033" , Litter_2="#00AED7",
             Sow_2="#00AED7", Litter_3="#FD9347", Sow_3="#FD9347", 
             Litter_4="#D55E00" , Sow_4="#D55E00", Litter_5="#FF3300", 
             Sow_5="#FF3300" , Litter_6="#3399CC" , Sow_6="#3399CC", 
             Litter_7="#FFCC00" , Sow_7="#FFCC00" , Litter_8="#006600", Sow_8="#006600", 
             Litter_9="#00CCFF", Sow_9="#00CCFF", Litter_10="#99CC00", Sow_10="#99CC00",  Actinobacteriota= "grey", Bacteroidota= "grey", Campilobacterota= "grey", Cyanobacteria= "grey",Desulfobacterota= "grey",
             Elusimicrobiota= "grey", Euryarchaeota= "grey", Firmicutes= "grey",
             Fusobacteriota= "grey", Patescibacteria= "grey", 
             Planctomycetota= "grey", Proteobacteria= "grey", Spirochaetota= "grey", 
             Synergistota= "grey", Thermoplasmatota= "grey", 
             Verrucomicrobiota= "grey")


#create a chord diagram but without labeling 
chordDiagram(data, grid.col = col, annotationTrack = "grid", preAllocateTracks = 1)

#add the labels and axis
circos.trackPlotRegion(track.index = 2, panel.fun = function(x, y) {
  xlim = get.cell.meta.data("xlim")
  ylim = get.cell.meta.data("ylim")
  sector.name = get.cell.meta.data("sector.index")
  
#print labels 
circos.text(mean(xlim), ylim[1] + 2.5, sector.name, 
              facing = "clockwise", niceFacing = TRUE, adj = c(0, 0.5), cex=0.6)
  
#print axis
circos.axis(h = "top", labels.cex = 0.5, major.tick.percentage = 0.2, 
              sector.index = sector.name, track.index = 2)
}, bg.border = NA)

```




